The effect of linking assumptions and number of response options on inferred scalar implicature rate

### 2018-05-08 writeup.Rmd
updated methods section

Consider a setup in which a listener is presented a card with a depiction of either one or two animals (see Figure 1 for an example). As in a standard truth-value judgment task, the listener then observes an underinformative utterance about this card (e.g., 'There is a cat or a dog on the card') and is asked to provide a judgment on a scale with 2, 3, 4, or 5 response options, with endpoints 'wrong' and 'right'.[^1] In the binary case, this reproduces the standard truth-value judgment task. Figure 1 exemplifies (some of) the researcher's options for grouping responses. Under what we will call the 'Strong link' assumption, only the negative endpoint of the scale is interpreted as evidence for a scalar implicature having been computed. Under the 'Weak link' assumption, in contrast, any response that does not correspond to the positive endpoint of the scale is interpreted as evidence for a scalar implicature having been computed. Intermediate grouping schemes are also possible, but these are the ones we will consider here. Note that for the binary case, the Weak and Strong link return the same categorization scheme, but for any number of response options greater than 2, the Weak and Strong link can in principle lead to differences in inferences about implicature rate.

@@

Participants played an online card game in which they were asked to judge descriptions of the contents of cards. Different groups of participants were presented with different numbers of response options. On critical trials, participants were presented with descriptions for the cards that typically result in exhaustive or scalar implicatures. We categorized their responses on such trials according to the "weak" and the "strong" link introduced above, and tested whether the number of response options and the linking assumptions led to different conclusions about the rate of computed implicatures in the experimental task.

@@

200 participants were recruited using Amazon Mechanical Turk. They optionally provided demographic information at the end of the study. Mean age of the participnats was `r round(mean_age)`. We also asked participants if they had any prior training in logic. `r logicTraining[[2]]` participants reported that they had while `r logicTraining[[1]]` had no prior training in logic. No participant was excluded from the final analysis.

### Materials and Procedure

The study was administered online through Amazon Mechanical Turk. Participants were first introduced to the set of cards we used in the study (Figure \@ref(fig:stimuli)). Each card had pictures of one or two animals. Animals were chosen from the following set: cat, dog, and elephant. Then participants were introduced to a blindfolded fictional character called Bob. Bob was blindfolded to control for violations of ignorance expectations with disjunction. Participants were told that Bob is going to guess what animals are on the card. We asked participants to let us know whether Bob's guess is wrong or right. In each trial, participants saw a card as well as a sentence representing Bob's guess. For example, they saw a card with a cat and read the sentence "There is a cat on the card." Depending on the task participants were assigned to, they had to choose between two (binary task), three (ternary task), four (quaternary task), or five (quinary task) response options. The study ended after 24 trials. You can access and view the study using the paper's [online repository](https://github.com/thegricean/si-paradigms/).

### Design

@@

The study had two main manipulaitons within participants: the type of card and the type of guess. There were two types of cards. Cards with only one animal on them and cards with two animals. There were three types of guesses: simple (e.g. *There is a cat*), conjunctive (e.g. *There is a cat and a dog*), and disjunctive (e.g. *There is a cat or a dog*). In each trial, the animal labels used in the guess and the animal images on the card may have no overlap (e.g. Image: cat, Guess: *There is an elephant*), a partial overlap (e.g. Image: cat, Guess: *There is a cat or a dog*), or a total overlap (e.g. Image: cat and dog, Guess: *There is a cat or a dog*). Crossing the number of animals on the card, the type of guess, and the overlap between the guess and the card results in 12 different possible trial types. We chose 8 trial types, balancing the number of one-animal vs. two-animal cards, simple vs. connective guesses, and expected true vs. false trials. Three trials were randomly selected from each of the 8 trial-types, for a total of 24 trials. The order of these 24 trials was randomized as well.

Participants could derive implicatures in two trial types. First, the trial type in which two animals were present on the card (e.g. cat and dog) but Bob guessed only one of them (e.g. "there is a cat"). Such trials can have a literal interpretation (cat and possibly more) or an exhaustive interpretation (only cat). We refer to them as "exhaustive". The second trial type with implicatures was the one in which two animals were on the card (e.g. cat and dog) and Bob used a disjunciton (e.g. cat or dog). These trials can have a literal (inclusive) interpretation (e.g. cat or dog or both), or an exclusive interpretation (e.g. cat or dog, not both). We refer to these trials as "scalar". The following four trial types were used as experimental control: two trial types where there was no overlap between the guess (e.g. elephant) and the animal(s) on the card (e.g. cat, cat and dog); and two trial types where the animal(s) on the card were correctly guessed. For example, if there was only a cat on the card, Bob said "there is a cat" and if there was a cat and a dog, Bob said "there is a cat and a dog". Since the fictional character was blindfolded and did not see the outcome of the game, the ignorance inference commonly associated with disjunction was already common ground in the experimental setting. If the character was seeing the cards or knew what was on them, a disjunction would have violated the expectation that the speaker does not know which alternative actually holds. Our study controls for the possibile effect of ignorance violations on exclusivity and exhaustive inferences.

The study also had a between participant manipulation of the number of response options in the forced choice task. Participants were randomly assigned to one of four different conditions or tasks. The tasks differed with respect to the number of response options: binary (wrong vs. right), ternary (wrong, neither, right), quaternary (wrong, kinda wrong, kinda right, right), and quinary (wrong, kinda wrong, neither, kinda right, right). We wanted to see if the number of response options in the forced choice task would affect our estimate of the task's "implicature rate".

@@

For our formal analysis, we used a bayesian binomial mixed effects model using the R packge "brms" [@burkner2016brms]. The model had the fixed effects of response type (binary, ternary, quaternary, quinary), linking assumption (strong vs. weak), and trial type (exhaustive vs. scalar), as well as their two way and three way interactions. We also included random intercepts and slopes for items (cards) and participants. Following @barr2013random, we used the maximal random effects structure by including random slopes for all our fixed effects and their interactions. Since the number of response options was a between participant variable we did not include random slopes of response options for participants. Four chains converged after 2000 iterations each (warmup = 1000). Table \@ref(tab:modeltable) summarizes the mean parameter estimates and their 95% credible intervals. $\hat{R}=1$ for all estimated parameters. All the analytical decisions described here were pre-registered[^2].

@@

[^2]: You can access our pre-registration at [https://aspredicted.org/tq3sz.pdf](https://aspredicted.org/tq3sz.pdf)

@@

@@

**XXX The following is entirely new:** However, this way of understanding behaivor in truth value jugdment tasks also raises new analytic and empirical questions which we must leave to future work. For instance, given the experimental results reported here, what is the range of plausible values for, e.g., the single probability threshold $\theta$, which separates "Right" from "Wrong" utterances in the binary condition? Bayesian statistical methods provide us the means for estimating the value of $\theta$ in our RSA-based model, by allowing us to recover a posterior distribution of possible values of $\theta$ given an assumption of a uniform prior and our empirically-observed data, i.e. the actual rate of response of "Right" in each trial of the binary condition of the experiment. We may also ask, for example, whether the speaker probability threshold for an utterance being "Right" is relatively constant across conditions or changes systematically with the introduction of more response options. Again, Bayesian data analysis allows us to explore this question by investigating the extent to which the inferred posterior distribution of, e.g., possible values of the threshold for "Right" in the binary condition of the experiment differs from the inferred posterior distribution of possible values of the threshold for "Right" in the ternary, quatenary, and quinary conditions.

>>>>

Consider a setup in which a listener is presented a card with a depiction of either one or two animals (see Figure \@ref(fig:linkvisualization) for an example). As in a standard truth-value judgment task, the listener then observes an underinformative utterance about this card (e.g., 'There is a cat or a dog on the card') and is asked to provide a judgment on a scale with 2, 3, 4, or 5 response options, with endpoints 'wrong' and 'right'.[^1] In the binary case, this reproduces the standard truth-value judgment task. Figure \@ref(fig:linkvisualization) exemplifies (some of) the researcher's options for grouping responses. Under what we will call the 'Strong link' assumption, only the negative endpoint of the scale is interpreted as evidence for a scalar implicature having been computed. Under the 'Weak link' assumption, in contrast, any response that does not correspond to the positive endpoint of the scale is interpreted as evidence for a scalar implicature having been computed. Intermediate grouping schemes are also possible, but these are the ones we will consider here. Note that for the binary case, the Weak and Strong link return the same categorization scheme, but for any number of response options greater than 2, the Weak and Strong link can in principle lead to differences in inferences about implicature rate.

@@

Participants played an online card game in which they were asked to judge descriptions of the contents of cards. Different groups of participants were presented with different numbers of response options. On critical trials, participants were presented with descriptions for the cards that typically result in exhaustivity implicatures ("There is a cat on the card" when there was a cat and a dog) or scalar implicatures ("There is a cat or a dog on the card" when there was a cat and a dog). We categorized their responses on such trials according to the weak and the strong link introduced above, and tested whether the number of response options and the linking assumptions led to different conclusions about the rate of computed implicatures in the experimental task.

@@

200 participants were recruited via Amazon Mechanical Turk. They optionally provided demographic information at the end of the study. Participants' mean age was `r round(mean_age)`. We also asked participants if they had any prior training in logic. `r logicTraining[[2]]` participants reported that they did, while `r logicTraining[[1]]` had no prior training in logic. **JD: Do we do any analysis of logic training???** All participants' data was included in the final analysis.

### Materials and procedure

The study was administered online through Amazon Mechanical Turk.[^2] Participants were first introduced to the set of cards we used in the study (Figure \@ref(fig:stimuli)). Each card depicted one or two animals, where an animal could be either a cat, a dog, or an elephant. Then participants were introduced to a blindfolded fictional character called Bob. Bob was blindfolded to avoid violations of ignorance expectations associated with the use of disjunction [@???] **JD: Masoud, can you insert citation pls? Geurts, Sauerland?**. Participants were told that Bob would guess the contents of the cards and their task was to indicate whether Bob's guess was wrong or right. On each trial, participants saw a card and a sentence representing Bob's guess. For example, they saw a card with a cat and read the sentence "There is a cat on the card." They then provided an assessment of Bob's guess.  The study ended after 24 trials.

[^2]: The experiment can be viewed [here](https://github.com/thegricean/si-paradigms/). **JD: Masoud, please turn this into a working link to the actual experiment**

@@

Two factors were manipulated within participants: card type and  guess type. There were two types of cards, cards with only one animal on them and cards with two animals. There were three types of guesses: simple (e.g. *There is a cat*), conjunctive (e.g. *There is a cat and a dog*), and disjunctive (e.g. *There is a cat or a dog*). Crossing card type and guess type yielded trials of varying theoretical interest (see Figure \@ref(fig:trials)): critical underinformative trials that were likely to elicit pragmatic  inferences  (either scalar or exhaustive) and  control trials that were either unambiguously true  or false. Each trial type occurred three times with randomly sampled animals and utterances that satisfied the constraint of the trial type. Trial order was randomized.

```{r trials, fig.asp=0.35, fig.cap = "Trial types (critical and control). Headers indicate utterance types. Rows indicate card types. Critical trials are marked in bold."}
 trials_img <- jpeg::readJPEG("writeup_files/figures/trialtypes.jpeg")
 grid::grid.raster(trials_img)
 ```

On critical trials, participants could derive implicatures in two ways. First, on trials on which two animals were present on the card (e.g., cat and dog) but Bob guessed only one of them (e.g. "There is a cat on the card"), the utterance could have a literal interpretation ("There is a cat and possibly another animal on the card") or an exhaustive interpretation ("There is only a cat on the card"). We refer to these trials as "exhaustive". Second, on trials on which two animals were on the card (e.g., a cat and a dog) and Bob used a disjunciton (e.g., "There is a cat or a dog on the card"), the utterance could have the literal, inclusive, interpretation, or a pragmatic, exclusive interpretation. We refer to these trials as "scalar".

In order to assess the effect of the number of response options on implicature rate, we manipulated number of response options in the forced choice task between participants. We refer to the choice conditions as 'binary' (options: *wrong*, *right*), 'ternary' (options: *wrong*, *neither*, *right*), 'quaternary' (options: *wrong*, *kinda wrong*, *kinda right*, *right*), and 'quinary' (*wrong*, *kinda wrong*, *neither*, *kinda right*, *right*). Thus, the endpoint labels always remained the same. If there was an uneven number of response options, the central option was *neither*.

@@

For our formal analysis, we used a bayesian binomial mixed effects model using the R packge "brms" [@burkner2016brms]. The model had the fixed effects of response type (binary, ternary, quaternary, quinary), linking assumption (strong vs. weak), and trial type (exhaustive vs. scalar), as well as their two way and three way interactions. We also included random intercepts and slopes for items (cards) and participants. Following @barr2013random, we used the maximal random effects structure by including random slopes for all our fixed effects and their interactions. Since the number of response options was a between participant variable we did not include random slopes of response options for participants. Four chains converged after 2000 iterations each (warmup = 1000). Table \@ref(tab:modeltable) summarizes the mean parameter estimates and their 95% credible intervals. $\hat{R}=1$ for all estimated parameters. All the analytical decisions described here were pre-registered[^3].

@@

[^3]: You can access our pre-registration at [https://aspredicted.org/tq3sz.pdf](https://aspredicted.org/tq3sz.pdf)

@@

**JD: moved from Methods section: Since the fictional character was blindfolded and did not see the outcome of the game, the ignorance inference commonly associated with disjunction was already common ground in the experimental setting. If the character was seeing the cards or knew what was on them, a disjunction would have violated the expectation that the speaker does not know which alternative actually holds. Our study controls for the possibile effect of ignorance violations on exclusivity and exhaustive inferences. **

@@

**XXX The following is entirely new:** However, this way of understanding behaivor in truth value jugdment tasks also raises new analytic and empirical questions which we must leave to future work. For instance, given the experimental results reported here, what is the range of plausible values for, e.g., the single probability threshold $\theta$, which separates "Right" from "Wrong" utterances in the binary condition? Bayesian statistical methods provide us the means for estimating the value of $\theta$ in our RSA-based model, by allowing us to recover a posterior distribution of possible values of $\theta$ given an assumption of a uniform prior and our empirically-observed data, i.e. the actual rate of response of "Right" On each trial of the binary condition of the experiment. We may also ask, for example, whether the speaker probability threshold for an utterance being "Right" is relatively constant across conditions or changes systematically with the introduction of more response options. Again, Bayesian data analysis allows us to explore this question by investigating the extent to which the inferred posterior distribution of, e.g., possible values of the threshold for "Right" in the binary condition of the experiment differs from the inferred posterior distribution of possible values of the threshold for "Right" in the ternary, quatenary, and quinary conditions.
